Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2024-12-22 ResponseLetter.tex

\textcolor{blue}{Thank you for your comment. We have made the change from 'cancer cells(' to 'cancer cells ('. change 25}

- results: eqn 6 has been pointed out by Orr $\&$ Unckless 2008 and Martin et al 2013

\textcolor{blue}{Thank you for your comment. We have mentioned that the results has been obtained previously by Orr $\&$ Unckless 2008 and Martin et al 2013. change 42}

- results: much of the discussion of the results (eg below eqn 7) is about the probability of rescue, but the equations all give the inverse of this, critical tumour sizes. I find it a little annoying to keep converting between them, and presenting the probabilities of rescue would also help compare with previous papers. Either way is fine but perhaps something to think about.

\textcolor{blue}{Thank you for your comment. We thought about this and we think that tumor size is critical for the interpretation of our model}

- results: I think the rescue times given in the main text (eqn 8) could be explained a little more. For small tumours, this is the 2-step version of the 1-step rescue time given in Orr $\&$ Unckless 2014 (the expectation of their eqn 18), $-1/r$. I.e., the faster the population declines the earlier rescue occurs (conditioned on it occurring), because that is when most mutations happen. For larger tumours this is just a deterministic constant population size result, where the successful mutation is produced at a constant rate.

\textcolor{blue}{Thank you for your comment. We have added comments that better explain the mean rescue time in the case of small and large tumors. change 46}

%- results: why doesn't standing genetic variance affect the mean rescue time? This might have something to do with what I just mentioned (successful mutants tend to arise early) and more clues might be found in Orr $\&$ Unckless 2014, who also compare times with and without standing genetic variance.

- results: 'aneuploidy promotes evolutionary rescue after $1/\Delta_a$' $\rightarrow$ 'aneuploidy promotes evolutionary rescue after $1/\Delta_s$'

\textcolor{blue}{Thank you for your comment. We have made the change from $1/\Delta_a$ to $1/\Delta_s$. change 37}

- results: in eqn 9 the variable $p_m$ is used, but it is not defined in the main text

\textcolor{blue}{Thank you for your comment. We have added the definition of $p_m$ below equation 9. change 26}

- results: the final term of the recurrence time for small tumours is equivalent to the return time of Orr $\&$ Unckless 2014 with standing genetic variation (eqn 16), but different from their return time with new mutation (eqn 23). Why?

\textcolor{blue}{Thank you for your comment. The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
\begin{align*}
t_{return}&\sim\frac{1}{s-r}\log\left[\frac{2N_0s(s-r)}{r}\right]\\
&\sim \frac{1}{s-r}\log[2N_0(s-r)]+\frac{1}{s-r}\log\left(\frac{s}{r}\right)
\end{align*}
Additionally, when $|s-r|<r$ we have:
\begin{equation*}
\log s \approx \log r +\frac{s-r}{r}
\end{equation*}
which can be rewritten as:
\begin{align*}
\frac{1}{s-r}\log\left(\frac{s}{r}\right)\approx \frac{1}{r}
\end{align*}
As a result, the return time with new mutation can be written as:
\begin{equation*}
t_{return}\sim \frac{1}{r}+\frac{1}{s-r}\log[2N_0(s-r)]
\end{equation*}
which is the one-step equivalent to our recurrence time for small tumors.
} 
- results: you mention that drugs increasing the death rate of sensitive cells delays recurrence, but given that mutations happen during division, it might be worth mentioning that drugs that decrease the division rate of sensitive cells by the same amount delay recurrence even more (holding mutant division rate constant - which is maybe why you didn't mention this for your parameterization, but seems like a worthwhile general point?).

\textcolor{blue}{Thank you for your comment. For large tumors, decreasing the division rate of sensitive cells will lead to a decrease in the cancer recurrence time when conditioned on evolutionary rescue. We have added this comment to the text. change 49}

- results: why doesn't standing genetic variance affect the mean rescue time? This might have something to do with what I just mentioned (successful mutants tend to arise early) and more clues might be found in Orr $\&$ Unckless 2014, who also compare times with and without standing genetic variance.

\textcolor{blue}{Thank you for your comment. Standing genetic variation does not have a significant impact on mean rescue time because the percentage of cell which have the beneficial aneuploidy is only $0.14\%$.}

- discussion: I find the discussion of the two advantages of aneuploidy for rescue a little confusing. Yes, aneuploidy delays population extinction and yes it increases the population size at a given time, but the important fact is that it increases the cumulative number of mutants that are created, which is the expression you give at the end of the sentence. Perhaps it would help to insert an additional sentence before the expression, like "Together, this increases the cumulative number of mutants that arise."

\textcolor{blue}{Thank you for your comment. We have added a sentence before the expression which states that aneuploidy increases the cumulant number of mutants that arise. change 39}

- discussion: the last sentence mentions that some cancer therapies are designed to increase aneuploidy, which felt like a new and important piece of information for me. Some discussion of this in the intro could be helpful as motivation for this work.

\textcolor{blue}{Thank you for your comment. In the last sentence of "Discussion" we state that certain therapies increase the rate of aneuploidy which has been shown by Lee et al. (2016) which is referenced in the "Introduction" section as a motivation for our research.}

- discussion: there are no citations of evolutionary rescue theory in the discussion, despite most of the results already existing or having very close parallels in the rescue literature. I think it would be nice to build more connection between rescue theory and cancer here.

\textcolor{blue}{Daniel ?.}

- table 1: what are the asterisks in the name column?

\textcolor{blue}{Thank you for your comment. We have added the definition of asterisk in the table. change 1}

- fig 2: when aneuploids are tolerant rescue can also happen indirectly (caption incorrect for panel B)

\textcolor{blue}{Thank you for your comment. We have made the change such that it says "When aneuploid cells are tolerant ($\Delta_a<0$) evolutionary rescue can occur indirectly, but direct mutation is the most likely path for evolutionary rescue". change 24}

- fig 4: give eqn $\#$ of exact formula (this goes for all formulas used in all figures)

\textcolor{blue}{Thank you for your comment. We added the references to the equation numbers for the formulas of $N_a^*$ and $N_m^*$. change 43}

- fig 4: $\lambda_m$ value given twice

\textcolor{blue}{Thank you for your comment. We have made the change from $\lambda_m$ to $\lambda_a$ in the first instance of $\lambda_m$. change 18}

- appendix A: 'decline are significant'$\rightarrow$'decline is significant'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 19}

- appendix A: in eqn A10, the $p_a$ and $p_m$ terms in the denominator seem to disappear in the second line (they will disappear in the third line anyway, given small mutation, but I think they should technically be in the second line)

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 20}

- appendix A: and in the third line of eqn A10 the $p_m$ in the numerator disappears (again, this doesn't matter for the result, eqn A11, but is an error and is required for the approximation to get from A10 to A11)

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 21}

- appendix C: $w_t\rightarrow s_t$

\textcolor{blue}{Thank you for your comment. We have made the requested change from $w_t$ to $s_t$. change 7}

- appendix C: I think the approximation in eqn C1 makes more sense by taking the Taylor series of the exponential term with respect to $\Delta_st$, rather than with respect to $t$ alone. For instance, you write $t + O(t^2)$, but the higher order terms are only smaller than t when $t\ll1$.

\textcolor{blue}{Thank you for your comment. We have changed $\mathcal{O}\left(t^2\right)$ to $\mathcal{O}\left(\Delta_st^2\right)$. change 22}

- appendix C: grammar mistakes in sentence below eqn C3

\textcolor{blue}{Thank you for your comment. We have corrected the grammatical mistakes. change 8}

- appendix C: I think it would be helpful to define $N_m^*$ and $N_a^*$ again in the appendix before using them

\textcolor{blue}{Thank you for your comment. We have added the definitions of $N_m^*$ and $N_a^*$ in Appendix C. change 38}

- appendix C: I would like to hear the justification for the initial expression of $\tau_a$ (integral of exponential) when $N>N_m^*$

\textcolor{blue}{Thank you for your comment. The justification is the following: 
\begin{align*}
\tau_a&=\int_0^\infty tf(t)dt=\int_0^\infty t\frac{d}{dt}F(t)dt=-\int_0^\infty t\frac{d}{dt}S(t)dt\\
&=-\left[tS(t)\right]_0^\infty+\int_0^\infty S(t)dt\\
&=\int_0^\infty S(t)dt,
\end{align*}
where f(t), F(t) and S(t) are the probability density function, cumulative distribution function and survival function.
In our case, $S(t)=\exp\left(-\int_0^t R_a(z)dz\right)$. Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$. I have corrected the following mistake:
\begin{equation*}
\tau_a=\int_0^\infty \e^{-\int_0^\tau R_a(t)\d t}\d\tau
\end{equation*}
instead of:
\begin{equation*}
\tau_a=\int_0^\infty \e^{-R_a(\tau)}\d\tau.
\end{equation*}
change 23
}

- appendix C: similarly, I think it would be nice to hear some justification for the approximations in eqn C9, ie, can ignore direct rescue and rescue via aneuploidy very unlikely

\textcolor{blue}{Thank you for your comment. We know that
\begin{align*}
\tau_a=\int_0^\infty t\left(v\lambda_ap_ma_t+v\lambda_sp_ms_t\right)\frac{\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] }{1-\e^{-Np_s}}\d t.
\end{align*}
Additionally, if $N\ll N_a^*$ then the probability distribution of evolutionary rescue time is skewed toward small values (i.e. evolutionary rescue occurs conditioned on it happening). As a result, we have: 
\begin{align*}
\frac{\e^{\Delta_s t}-1}{\Delta_s}&=\frac{1+\Delta_s t+\mathcal{O}\left(\left(\Delta_st\right)^2\right)-1}{\Delta_s}=t+\mathcal{O}\left(\Delta_st^2\right)\\
\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+\mathcal{O}\left(\left(\Delta_at\right)^2\right)-1}{\Delta_a}=t+\mathcal{O}\left(\Delta_at^2\right)
\end{align*}
We observe that the first term in the exponential can be approximated to be zero and the second term as -$v\lambda_sNp_mt$ where $t\ll1$. As a result, we can approximate:
\begin{equation*}
\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] \sim 1
\end{equation*}
Additionally, since $N\ll N_a^*\ll N_m^*$, evolutionary rescue is more likely to occur through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$. As a result,
\begin{equation*}
\tau_a\sim\int_0^\infty \frac{tv\lambda_ap_ma_t}{1-\e^{-Np_s}}\d t.
\end{equation*}
}

- appendix D: the fact that lineages conditioned to survive drift reach N earlier is nicely discussed in Orr $\&$ Unckless 2014, who connect it back to classic work on selective sweeps by Maynard Smith $\&$ Haigh 1974

\textcolor{blue}{Thank you for your comment. We have added citations for Orr $\&$ Unckless 2014 and Maynard Smith $\&$ Haigh 1974. change 35}

- appendix D: 'it accurate' $\rightarrow$ it is accurate'

\textcolor{blue}{Thank you for your comment. We have made the change from 'it accurate' to 'it is accurate'. change 9}

- appendix F: 'Figure Figure' $\rightarrow$ 'Figure'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 10}

- appendix F: 'our approximation are'$\rightarrow$'our approximations are'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 11}

- appendix F: 'From eq. (F2) we obtain'$\rightarrow$'From eq. (F1) we obtain'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 12}

- fig S2: 'the leads'$\rightarrow$ 'that leads', 'is in agreement'$\rightarrow$'are in agreement', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 13}

- fig S3: 'number of mutation'$\rightarrow$'number of mutations', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 14}

- fig S5: $\mu_a$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 15}

- fig S6: 'the leads'$\rightarrow$'that leads', 'lines, right'$\rightarrow$'lines,', 'is in agreement'$\rightarrow$'are in agreement', 'lines represents'$\rightarrow$'lines represent', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 16}

- fig S9: 'the leads'$\rightarrow$'that leads', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 17}

Reviewer $\#3$ :

Summary

Stana et al. study the effects of aneuploidy (mutations that give rise to atypical karyotypes) on the survival of a tumor- a population of malignant cells- in presence of an intervention, here, an anti-cancer drug. While the role of aneuploidy in development of drug-resistance has been acknowledged in the field of evolutionary oncology, the precise mechanism by which it incurs drug-resistance is not clear.

This manuscript is about understanding the evolutionary role of aneuploidy in drug-resistance in cancer. The authors model tumor evolutionary dynamics under treatment as a multi-type branching process, where the types correspond to three genotypes: sensitive to therapy, mutant (resistant), and aneuploid. The key assumption is that aneuploid cells are less-sensitive than sensitive cells to therapy. They parameterized their model by the pre/post treatment cost of aneuploidy, mutation rates and birth/death rates for each genotype.

Stana et al. derive equations for relevant evolutionary quantities under multiple parameter regimes. The evolutionary quantities include the probability of evolutionary rescue, and time to recurrence. They further investigate the effects of standing variation, and density-dependent growth. A central finding is that aneuploidy can directly or indirectly result in evolutionary rescue. In the latter, it delays tumor extinction and provides a window of opportunity for mutations to rescue the population.

Overall, the paper proposes a rigorous evolutionary approach to understand the effects of aneuploidy in development of drug-resistance in cancer. The authors verify the accuracy of their derivations via extensive stochastic simulations. The paper is well written and the ideas are presented in a cohesive and rational manner. The accompanying python implementation appears sufficient (see below) to reproduce the results. The problem is interesting and subject of active research in evolutionary oncology, specially development of drug-resistance subpopulations (PMID: 39223250) and new avenues of targeting aneuploid cells (PMID: 38992122).

Major points

- The importance of time-resolved measurements of clonal dynamics have been recognized and more such datasets are becoming available (e.g., PMID: 34163070, PMID: 38987605). The simulation studies in this manuscript use a parameter regime largely based on a melanoma cell-line (A375) and treatment with a specific BRAF-inhibitor (vemurafenib). The relevance of the paper would be improved if consistent results can be shown in other datasets, including the citations above.

Minor points

- An exciting finding of the paper is that aneuploidy can act as an evolutionary stepping stone (Figure 4). However, in the discussion, you point out that only in a small parameter regime this would be the case, and in most cases, aneuploidy either directly results in evolutionary rescue or does not play a role. If this were the case, wouldn't you expect that a majority of cancer cells in a tumor to either have clonal/truncal aneuploidies or have no clonal aneuploidies at all? Given the prevalence of aneuploidies in patient cohorts, how can we interpret the restricted parameter regime that allows it?

\textcolor{blue}{Thank you for your comment. In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs. As a consequence, we catalogued all other forms of aneuploids as sensitive.}

- Code Base
The python scripts used for simulation studies are provided in the supplementary package, however the github link is not accessible.
It would improve the reproducibility of the manuscript if the code to reproduce individual figures is provided.
Documenting the code as well as adding a short tutorial will improve the reusability of the scripts.

- In your model, mutations appear to only confer resistance, however, most mutations are deleterious. How easy is it to incorporate the effects of deleterious mutations in your model? This should be addressed in the discussion.

\textcolor{blue}{Thank you for your comment. We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population. As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population. I have added this comment in the "Parametrisation" section change 50}

- You report the exact parameters used in most figures. Sometimes it is difficult to follow why a specific value for a parameter was chosen. It would improve the readability if you included a table where the variation in the parameter used is noted and explained.

\textcolor{blue}{Thank you for your comment. The variation in our parameters in noted in Table 1 and is obtained from the papers which are cited there.}

Associate Editor Comments:

The paper sits in an odd place, as noted by reviewer $\#2$. Existing theory already explores evolutionary rescue by one and two step mutations, and it is hard to see the connection between the current theory and the existing theory. Which of the results are new? It is important that the reader knows what is novel. If similar results are derived in a different way, please explain the connection. When the quantity considered is different (e.g., timing of certain events), please alert the reader and maybe explain why these quantities are particularly relevant in the context of cancer. Reviewer $\#2$ has many helpful suggests along these lines.

Another oddity is that the theory developed applies much more broadly to evolutionary rescue scenarios, but the presentation makes it seems that the results only apply to cancer. Similarly, the initial facilitating mutation could be any type of mutation, not just aneuploidy.

A sufficient revision must, in my view, properly connect this work to the existing theoretical literature on evolutionary rescue (as noted by reviewer $\#2$) and to the broader empirical literature on cancer about multi-step drug resistance (as noted by reviewer $\#3$).

Minor comments
Introduction:
* 10 millions people die: add globally

### 2024-12-22 ResponseLetter.tex

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 30}

* mathematically equivalent: this is incorrect because fitness-valley crossing papers typically ignore population size dynamics, whereas evolutionary rescue is explicitly set in the context of a declining population size.

\textcolor{blue}{Thank you for your commnet. We have changed the text such that it states that evolutionary rescue in two steps is similar to the problem of crossing fitness valley. change 6}

* how multiple mutations contribute: several important papers on this subject are referenced later but not in this paragraph. Please revise, reviewing the existing theory and saying how the current theoretical work differs.

Methods:
* The methods only describe the simulations, and the results focus on analytical results that don't follow from the methods. I suggest expanding the "Methods" to discuss the general approach, model structure, and quantities derived (you can rename this as "Model") and then move to "Results".

\textcolor{blue}{Thank you for your commnet. The model is described in "Methods" under the subsection "Evolutionary model" and the analytic results follow directly from this model. Additionally, the quantities derived are stated in the paragraph before "Methods". }

* Delta would be better as "r" for the growth rate

\textcolor{blue}{Thank you for your commnet. The are about 300 mentions of $\Delta$ and we do not want to introduce any errors by changing $\Delta$ to $r$.}

* The use of the terms "tolerant" (for negative growth) and "resistant" (for positive growth) is not standard (e.g., see different for fungi used by Berman $\&$ Krysan 2020 Nature Reviews Microbiology). Maybe alert the reader that "tolerant" is used in different ways in the literature and is used here solely for notational convenience to refer to mutations that, on their own, do not allow growth? Or replace with "sub-critical", a term already used in this context?

\textcolor{blue}{Thank you for your commnet. We have added a short explanation for "tolerant" and that it has other meanings in the literature. change 47.}

* "antagonizing cell division" - The meaning of this isn't clear. Rephrase as "by extending the length of the cell cycle"?

\textcolor{blue}{Thank you for your comment. We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle". change 34}

Results:
* Figure 7 - Add a numerical solution to equation (D3)? This should make it clearer that the non-monotonicity of the black curve is an artefact of the simplifying assumptions in (D4) and not a problem with the approach. By the way, It wasn't clear to me what happened to the other large $\tau_a^r$ terms in the first portion of equation (D3) when approximating the equation into the form shown in (D4)? Could this be briefly clarified?

\textcolor{blue}{Thank you for your comment. We have changed Figure 7 by adding a green which represents the numerical solution of (D3).  In the first term of (D3) we have:
\begin{align*}
\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
\end{align*}
which is significantly smaller then the second term of (D3):
\begin{align*}
v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)}
\end{align*}
change 44}

Remus changes:

\textcolor{red}{$w_t\rightarrow s_t$. change r1}

\textcolor{red}{$t\rightarrow \tau$. change r2}

\textcolor{red}{
We have made the following changes: the rates of the inhomogenous Poisson process are:
\begin{align*}
r_1\left(t\right)&=v\lambda_ap_ma_{t},\\ 
r_2\left(t\right)&=v\lambda_sp_ms_{t},
\end{align*}
and the expected number of successful mutants created by direct mutation and via aneuploidy until time $t$ is given by:
\begin{align*}
M_a\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ 
M_m\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.
\end{align*}
change r3}

\textcolor{red}{We have made changes to Appendix C such that all the dummy variable used in integration are consistent. change r4}

\textcolor{red}{We have made changes to Figure S6, S7, S8, S9 by changing the subscripts of several variables from $w$ to $s$ . change r5}

\textcolor{red}{ We have made the following change: "If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as" $\rightarrow$
"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by". change r6}

\bibliographystyle{plain}
\bibliography{evo2022}
\end{document}
